---
title: "Data Processing"
subtitle: "How the BRFSS data were acquired, cleaned, harmonized, and explored"
author: "Boqian Mao"
date: today
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 8, fig.height = 5
)
```

```{r libraries}
library(tidyverse)
library(knitr)
library(DT)
library(plotly)
library(scales)
```

```{r load-data}
brfss <- readRDS("brfss_subset.rds")

base_all <- brfss |> filter(!is.na(fmd))
base_2024 <- base_all |> filter(year == 2024)

n_total_all <- nrow(brfss)
n_base_all  <- nrow(base_all)
n_base_2024 <- nrow(base_2024)
n_states_2024 <- dplyr::n_distinct(base_2024$state)
```

# Overview

This page documents the full data pipeline behind the main report. The analysis uses five consecutive years of CDC Behavioral Risk Factor Surveillance System (BRFSS) public-use survey data, then trims and harmonizes those files into a smaller analytic dataset focused on frequent mental distress (FMD), geography, demographic factors, and selected behavioral and psychosocial correlates.

The preparation code used to build the analytic dataset is available in the project repository at [`prep_brfss_subset.R`](https://github.com/Alfie0925/biostat620-final-project/blob/main/prep_brfss_subset.R). The rendered website reads the resulting analytic file from `brfss_subset.rds`.

# Data Background

This project uses multiple years of data from the Behavioral Risk Factor Surveillance System (BRFSS), a large national health survey conducted by the Centers for Disease Control and Prevention (CDC) in collaboration with U.S. states and participating territories. Established in 1984, BRFSS collects standardized, state-specific information on health-related risk behaviors, chronic conditions, health care access, and preventive service use among noninstitutionalized adults aged 18 years or older in the United States.

BRFSS data are based on self-reported telephone interviews and, since 2011, have been collected using both landline and cellular telephones. This dual-frame design improves population coverage and representativeness as telephone use patterns have shifted over time. Survey data are collected by state health departments following CDC protocols, then sent to the CDC for editing, weighting, and public release.

To improve representativeness, BRFSS uses iterative proportional fitting ("raking") to generate survey weights. This method adjusts the sample to better reflect the demographic makeup of each state using characteristics such as age, sex/gender, race and ethnicity, education, marital status, home ownership, geographic region, and telephone source.

Across years, BRFSS follows a common structure that includes a core questionnaire, optional modules, and state-added questions. The core questionnaire is administered across participating states, while optional modules may vary by year and location. By combining multiple annual BRFSS datasets, this study uses a large, well-established public health surveillance source while accounting for modest year-to-year differences in survey content and participation.

# Raw data acquisition

The raw input files are the annual BRFSS public-use SAS transport files released by the CDC. I worked from the 2020 through 2024 annual survey cycles and imported the `.XPT` files with `haven::read_xpt()`.

```{r raw-data-table}
raw_tbl <- tribble(
  ~Year, ~File, ~Source,
  2020, "LLCP2020.XPT", "[CDC 2020 annual data](https://www.cdc.gov/brfss/annual_data/annual_2020.html)",
  2021, "LLCP2021.XPT", "[CDC 2021 annual data](https://www.cdc.gov/brfss/annual_data/annual_2021.html)",
  2022, "LLCP2022.XPT", "[CDC 2022 annual data](https://www.cdc.gov/brfss/annual_data/annual_2022.html)",
  2023, "LLCP2023.XPT", "[CDC 2023 annual data](https://www.cdc.gov/brfss/annual_data/annual_2023.html)",
  2024, "LLCP2024.XPT", "[CDC 2024 annual data](https://www.cdc.gov/brfss/annual_data/annual_2024.html)"
)

kable(raw_tbl, escape = FALSE,
      caption = "Raw BRFSS annual survey files used to construct the analytic dataset.")
```

The broader BRFSS data portal is the CDC's [Annual Survey Data](https://www.cdc.gov/brfss/annual_data/annual_data.htm) page. Across these five years, the public-use files contain a little over 2.18 million respondent records before analytic filtering.

# Variables retained and harmonized

The main report focuses on frequent mental distress plus a set of geography, time, demographic, socioeconomic, behavioral, and psychosocial variables. Only those variables were carried forward into the analytic dataset.

```{r var-table}
var_tbl <- tribble(
  ~Variable, ~Description, ~Role,
  "_MENT14D",  "14-day mental-distress indicator (CDC derived)",            "Primary outcome (FMD)",
  "MENTHLTH",  "Number of poor-mental-health days in past 30",              "Secondary outcome",
  "_STATE",    "FIPS state code",                                           "Geography",
  "year",      "BRFSS cycle year (2020-2024)",                              "Time",
  "EXERANY2",  "Any physical activity in past 30 days",                     "Behavioral",
  "SDLONELY",  "How often felt lonely (optional module, 2023-2024)",        "Psychosocial",
  "EMTSUPRT",  "How often received needed emotional support",               "Psychosocial",
  "LSATISFY",  "Life satisfaction",                                         "Psychosocial",
  "GENHLTH",   "Self-rated general health",                                 "Health status",
  "_AGEG5YR",  "Age in 5-year bands",                                       "Demographic",
  "_SEX",      "Sex at birth",                                              "Demographic",
  "_INCOMG1",  "Income category, harmonized to 5 bins",                     "Socioeconomic",
  "EDUCA",     "Highest education completed",                               "Socioeconomic",
  "_BMI5CAT",  "BMI category (derived)",                                    "Health status",
  "MEDCOST1",  "Could not see a doctor due to cost",                        "Access to care",
  "EMPLOY1",   "Employment status",                                         "Socioeconomic",
  "MARITAL",   "Marital status",                                            "Socioeconomic",
  "_SMOKER3",  "Smoking status (derived)",                                  "Lifestyle",
  "DRNKANY6",  "Any alcohol in past 30 days",                               "Lifestyle"
)

kable(var_tbl, caption = "Variables retained from the raw BRFSS files.")
```

Three variables had to be explicitly harmonized across cycles before binding the files together:

- `_INCOMG` in 2020 was copied into `_INCOMG1` to match the 2021+ naming convention.
- `MEDCOST` in 2020 was copied into `MEDCOST1`.
- `DRNKANY5` in 2020-2021 was copied into `DRNKANY6` to align with the 2022+ variable name.

Income also needed a content harmonization, not just a rename. The 2020 BRFSS income field used a 5-level scheme, while later years split the top category into multiple higher-income bins. To keep the cross-year comparisons valid, I collapsed every cycle into a shared 5-level scale: `<$15k`, `$15-25k`, `$25-35k`, `$35-50k`, and `$50k+`.

# Cleaning and recoding

The preparation script followed a small set of consistent rules before saving the analytic file:

1. Read each yearly `.XPT` file and retain only the variables needed for the report.
2. Add a `year` column and standardize renamed variables so all years share one schema.
3. Remove U.S. territories by keeping only the 50 states plus the District of Columbia.
4. Map `_STATE` FIPS codes to full state names and two-letter abbreviations.
5. Recode BRFSS non-substantive responses to missing. For most categorical variables, codes such as 7 ("Don't know/Not sure"), 9 ("Refused"), and blank "Not asked or Missing" were treated as `NA`. For `MENTHLTH`, 77 and 99 were treated as `NA`, while 88 ("None") was recoded to 0 poor-mental-health days rather than missing.
6. Define frequent mental distress as `_MENT14D == 3`, matching the CDC's 14-day threshold.
7. Recode selected variables into labelled factors for plotting and tabulation.
8. Save the cleaned analytic dataset as `brfss_subset.rds`.

# Sample-size flow

```{r sample-flow}
flow_year <- base_all |>
  count(year, name = "n_valid_FMD")

n_lone_all <- sum(!is.na(base_all$loneliness))

flow <- tribble(
  ~Step, ~N,
  "All 2020-2024 respondents in the cleaned combined file",         n_total_all,
  "Base cohort with valid FMD (50 states + DC, all years)",         n_base_all,
  "2024 cycle only with valid FMD",                                 n_base_2024,
  "Subset with valid loneliness/support items (optional module)",   n_lone_all
)

kable(flow, format.args = list(big.mark = ","),
      caption = "Sample-size flow from combined file to analytic cohorts.")

kable(flow_year, format.args = list(big.mark = ","),
      col.names = c("Year", "N (valid FMD)"),
      caption = "Valid-FMD sample size by BRFSS cycle.")
```

The main report uses pairwise complete cases rather than listwise deletion. That means a respondent can still contribute to a figure even if they are missing an unrelated variable used somewhere else on the site.

# Exploratory checks

These exploratory summaries helped confirm that the cleaned file looked sensible before writing the final report.

## Records by year

```{r eda-year-counts}
year_counts <- brfss |>
  count(year, name = "n")

plot_ly(year_counts,
        x = ~year, y = ~n, type = "bar",
        marker = list(color = "#2980b9"),
        text = ~comma(n), textposition = "outside",
        hovertemplate = "Year %{x}<br>Rows: %{y:,}<extra></extra>") |>
  layout(
    title = list(text = "Exploratory check: analytic rows available by year", x = 0),
    xaxis = list(title = "BRFSS cycle", dtick = 1),
    yaxis = list(title = "Rows retained after cleaning"),
    showlegend = FALSE,
    margin = list(t = 60)
  )
```

## State sample sizes by year

```{r eda-state-sizes, fig.height=14, fig.width=10}
year_state_counts <- base_all |>
  count(year, state, name = "n")

state_order <- year_state_counts |>
  group_by(state) |>
  summarise(mean_n = mean(n), .groups = "drop") |>
  arrange(mean_n) |>
  pull(state)

year_state_counts |>
  mutate(state = factor(state, levels = state_order)) |>
  ggplot(aes(x = n, y = state, fill = n)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~year, ncol = 2, scales = "free_x") +
  scale_x_continuous(labels = comma) +
  scale_fill_gradient(low = "#d6eaf8", high = "#2471a3") +
  labs(
    title = "Exploratory check: state sample sizes with valid FMD by year",
    x = "Respondents with valid FMD",
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.y = element_text(size = 6),
    panel.grid.major.y = element_blank()
  )
```

This faceted bar chart shows how the valid-FMD sample size varies across states within each BRFSS cycle. The overall state ordering is based on average sample size across years, which makes it easier to compare how stable the state-level data coverage is from 2020 through 2024.

## Distribution of poor mental health days

```{r eda-mental-days}
mental_days_dist <- base_2024 |>
  filter(!is.na(mental_days)) |>
  count(mental_days) |>
  mutate(
    group = case_when(
      mental_days == 0  ~ "0 days",
      mental_days == 30 ~ "30 days",
      mental_days >= 14 ~ "14+ days",
      TRUE              ~ "1-13 days"
    ),
    group = factor(group, levels = c("0 days", "1-13 days", "14+ days", "30 days"))
  )

plot_ly(
  mental_days_dist,
  x = ~mental_days, y = ~n, color = ~group,
  colors = c("0 days" = "#27ae60", "1-13 days" = "#bdc3c7",
             "14+ days" = "#e67e22", "30 days" = "#c0392b"),
  type = "bar",
  hovertemplate = "%{x} days<br>Respondents: %{y:,}<extra></extra>"
) |>
  layout(
    title = list(text = "Exploratory check: poor mental health days in 2024", x = 0),
    xaxis = list(title = "Days of poor mental health in the last 30 days", dtick = 5),
    yaxis = list(title = "Respondents"),
    legend = list(title = list(text = "Category"), orientation = "h",
                  x = 0.5, xanchor = "center", y = -0.2),
    barmode = "stack",
    margin = list(t = 60, b = 80)
  )
```

This distribution is one reason the dichotomized FMD outcome is useful here: respondents cluster heavily at 0 days and again at the extreme upper tail, rather than forming a smooth continuous distribution.
